Determining sequence-dependent DNA oligonucleotide hybridization and dehybridization mechanisms using coarse-grained molecular simulation, Markov state models, and infrared spectroscopy

### 2021-05-13 manuscript.tex
Andy edits

%include hyperlinks to figures
%\documentclass{article}
\usepackage{graphicx}
\usepackage[%  
    colorlinks=true,
    pdfborder={0 0 0},
    linkcolor=blue,
    citecolor=black
]{hyperref}

@@

\begin{figure}[ht!]
	\begin{center}
        \includegraphics[width=1.0\textwidth]{Fig2.pdf}
	\end{center}
\end{figure}
\begin{figure}[ht!]
        \caption{Thermodynamic and kinetic predictions of the sequence-specific MSMs for AT-all, GC-end, GC-core, and GC-mix. (a) \textbf{Macrostates.} Schematic representation of the seven metastable macrostates occupied by one or more of the four sequences: fully hybridized state (H), two-base out-of-register shifted states in the 5$^\prime$ (5S2) or 3$^\prime$ (3S2) direction, four-base out-of-register shifted states in the 5$^\prime$ (5S4) or 3$^\prime$ (3S4) direction, and the fully dissociated state (D). The line drawings represent the 10-base self-complementary sequences, where red-to-blue contacts indicates (possible) WC base pairing and black indicates an unbound bases. Adjacent to each line drawing we provide representative molecular structures corresponding to that macrostate. (b) \textbf{Thermodynamics.} Histograms reporting the number of the $10^7$ total frames within the 1 ms of simulation trajectories observed to occupy each of the seven macrostates, corresponding to our numerical estimates of the equilibrium occupancy probabilities. Uncertainties are calculated across 100 MSMs using a Bayesian MSM estimation are reported for each bar and are very small compared to the total counts. (c) \textbf{Kinetic networks.} MSMs illustrating the kinetic network for each sequence. The orange circles correspond to the macrostates occupied by each sequence and are labeled by the macrostate codes reported in panel a. The radii of the circles is proportional to the logarithm of the equilibrium occupancy probabilities reported in panel b. Molecular renderings of an illustrative snapshot from the coarse-grained molecular simulations are provided next to each macrostate. The grey arrows between macrostates indicate the presence of a kinetic flux between this pair of states at equilibrium and the arrow thickness is proportional to the flux. (To avoid congesting the diagram, arrows are not reported for fluxes lower than $3\times10^{-6}$ \noteb{flux is in terms of $p_i*p_{ij}$ so I think it should still be unitless?}.) The numerical value overlaid on each arrow reports the conditional probability that a system occupying the macrostate at the start of the arrow at time $t$ will transition to the macrostate at the end of the arrow by time $(t+\tau)$, where $\tau$=1.2 ns is the lag time corresponding to a single time step of the MSM. Large orange circles correspond to thermodynamically favorable states and large grey arrows correspond to kinetically favorable transitions. (d) \textbf{Kinetic time scales.} Distribution of MSM implied time scales for each sequence. The leading implied time scale corresponds to the characteristic time scale for hybridization/dehybridization and is approximately equal for all systems since simulations were conducted at the same concentration and at the respective melting temperatures. The higher order implied time scales correspond to a spectrum of kinetic relaxations between the constituent macrostates in the MSM corresponding to shifted and/or frayed states.
    }
    \label{fig:allseq_table}
\end{figure}

@@

AT-all possesses the richest and most complex MSM of the four sequences by virtue of its repetitive nature, comprising a hybridized state (H), dissociated state (D), and four out-of-register shifted states (5S2, 3S2, 5S4, 3S4) (Fig.~\ref{fig:allseq_table}c). Analysis of the MSM transition probabilities reveal a critical role of the shifted states in mediating hybridization and dehybridization. Commencing from the dissociated state D, we observe approximately equal probabilities for transitions to each of the other five states, such that a transition to one of the out-of-register shifted states 5S2, 3S2, 5S4, or 3S4 is approximately 2.2 times more likely than a direct transition to the hybridized state H. Commencing from the hybridized state H, however, a direct transition to the dissociated state is approximately 1.2 times more likely than a transition to one of the two-base shifted states 5S2 or 3S2. Once in one of the four shifted states, the 5$^\prime$ vs. 3$^\prime$ overhang and degree of shifting play an important role in determining whether the duplex will transition to more shifted states, more aligned states, or completely dissociate. Transitions from more shifted states towards more aligned states (i.e., 5S4 $\rightarrow$ 5S2, 5S2 $\rightarrow$ H, 3S4 $\rightarrow$ 3S2, 3S2 $\rightarrow$ H) are approximately an order of magnitude more probable than the reverse transitions from more aligned states to more shifted states. The largest single transition probability from the four shifted states 5S2, 3S2, 5S4, and 3S4 is, however, back to the dissociated state D. Consistent with the higher destabilizing effect of 3$^\prime$ inert tails relative to 5$^\prime$ tails,~\citep{Doktycz1990ThermodynamicATGC, Dickman2012ThermodynamicDNAs, Michele2014EHybridization} the 3S4 $\rightarrow$ D transition probability is twice as large as the 5S4 $\rightarrow$ D, and the 3S2 $\rightarrow$ D is four times larger than the 5S2 $\rightarrow$ D. The transition probability from the 5S2 and 3S2 states back to the dissociated state D is equal to or greater than the transition probability to the hybridized state. A transition path theory analysis of the MSM reveals that 33\% of productive hybridization trajectories D $\dashrightarrow$ H (where the dashed arrow indicates the combination of both direct and indirect pathways) proceed through one or more out-of-register shifted states. Among these out-of-register pathways, the D $\dashrightarrow$ 5S2 $\dashrightarrow$ H transition is predicted to occur 57\% of the time. A mean first passage time (MFPT) analysis returns a MFPT for D $\dashrightarrow$ H of 3.0 $\mu$s and for H $\dashrightarrow$ D of 2.5 $\mu$s. \noteb{Can add additional analysis "When considering the mean MFPT between D and any state A (A = 5S2 $\cup$ 3S2 $\cup$ 5S4 $\cup$ 3S4 $\cup$ H), we observe the shorter time scales 1.3 $\mu$s for A $\rightarrow$ D 1.3 $\mu$s for D $\rightarrow$ A. These MFPTs reflect the frequency of partial hybridization events and are in close agreement with the leading implied time scale of $t_1$ = 1.5 $\mu$s reported in Fig.~\ref{fig:allseq_table}d" This works for AT-all, but for all sequences the leading timescale tends to be lower by up to 2x than H $\dashrightarrow$ D MFTPs. I suspect this is because these leading timescale takes longer to converge, but I'm not sure if we want to draw attention to this and/or report the leading timescale for each sequence since its exact meaning is not always clear)} As expected by the fact that the calculations are performed at the approximate melting temperature of the sequence, the MFPTs are approximately equal to one another.

@@

\noteb{and in good agreement with the leading implied time scale of $t_1$ = 1.4 $\mu$s reported in Fig.~\ref{fig:allseq_table}d.}

@@

\noteb{This MFPT shows the largest deviation from the leading timescale $t_1$ = 1.6 $\mu$s}

@@

\noteb{Again, these are nearly equal and commensurate with the dominant leading implied time scale of $t_1$ = 2.0 $\mu$s reported in Fig.~\ref{fig:allseq_table}d. }

@@

\subsection{\label{sec:stretch_comparison}Long-lived metastable states predicted by the MSM are resolved by T-jump IR}

\noteb{After running more low-temp simulations and looking at NN predictions for shifted states at low temps, it seems like the shifted equilibrium does not change too much (a relatively higher loss for GC-end than AT-all). I think the state lifetime argument becomes confusing since we (i) we are presenting simulation and experimental time scales on the same figure (ii) do not have an acceleration factor to readily convert these timescales (iii) do not have any evidence of mid-range relaxation that you might expect given this timescale spectrum. No need to mention any of this in the paper, but I think there is really a catch-22 with trying to detect these states with T-jump. The only process you can really resolve is going from shifted --> D since H--> shifted will be as slow as H-->D. If a state is unstable this process should happen at an intermediate timescale, but it contain such a small (less than 1\%) of the total population that it won't be picked up. If the metastable state is more stable then the population will be larger, but the process will occur near the same timescale as full dissociation and look the exact same.  So all of that is to say that a stretched exponential can very reasonably be attributed to shifted-fraying response, but it becomes more of a thermodynamic argument than kinetic.}

We hypothesized that we should be able to validate particular kinetic predictions of our sequence-dependent MSMs using T-jump IR spectroscopy. T-jump IR measurements commencing from low temperatures (at which the hybridized state dominates the equilibrium ensemble) apply a rapid step jump in temperature and probe the relaxation of the system to the dehybridized state. As discussed in Section \ref{sec:time}, the fast and slow IR responses correspond well to terminal base fraying and dissociation, respectively. Shifting dynamics are more difficult to distinguish given that shifting events (e.g. H $\rightarrow$ 5S2) are shown by our MSMs to occur on similar time scales as H $\rightarrow$ D and would therefore be blended into the slow response. However, we do expect some portion of low temperature duplexes to initialize in shifted states and alter fast fraying dynamics. Furthermore, we expect this effect to be proportional to the population of shifted states relative to intact hybridized states. At our MSM equilibrium distribution, we find that AT-all shifted states make up 24\% of the hybridized ensemble (all states excluding D) compared to only 0.3\% for GC-end shifted states. GC-core and GC-mix would not be expected to have any shifted states in the starting ensemble. Based on these predictions from our equilibrium MSMs, we hypothesized that the AT-all sequence, possessing significantly more heterogeneous state distribution after the T-jump, should be distinguishable from the GC-end, GC-core, and GC-mix sequences. We do not expect to be able to resolve the small differences between GC-end, GC-core, and GC-mix due to the similarity of the starting ensemble.

We present in Fig.~\ref{fig:T-jump_stretching} our T-jump IR t-HDVE difference spectra and corresponding normalized time traces at 1600 cm$^{-1}$ and 1660 cm$^{-1}$. The signal at 1600 cm$^{-1}$ corresponds to changes in A and T ring vibrations while the signal at 1660 cm$^{-1}$ contains contributions from G and T carbonyl vibrations. Each time trace was fitted to the sum of a stretched exponential and two exponentials $S(t) = A\exp{(-(t/\tau^{\mathrm{fast}})^{\beta})} + B\exp{(-t/\tau_{\mathrm{slow}})}+C\exp{(-t/\tau_{\mathrm{cool}})}$. The stretched exponential describes the relaxation process from 5 ns to 1 $\mu$s, and the two exponentials describe the signal increase from 1-320 $\mu$s and signal decay from re-hybridization induced by thermal relaxation back to the initial temperature. The fitting parameters $A$, $B$, and $C$ correspond to the relative amplitudes of the three kinetic responses and we attribute the stretch factor $\beta$ to the heterogeneity of fraying dynamics at the fast timescale. A testable prediction of our hypothesis is that the long-lived out-of-register shifted states in AT-all should result in a significantly smaller value for the fitted $\beta$ parameter relative to those for GC-end, GC-core, and GC-mix. This hypothesis was supported by the experimental time series at both 1600 cm$^{-1}$, where $\beta_\mathrm{AT-all}$ = 0.3 compared to $\beta_\mathrm{GC-end}$ = 0.6, $\beta_\mathrm{GC-core}$ = 0.7, and $\beta_\mathrm{GC-mix}$ = 0.6, and 1660 cm$^{-1}$, where $\beta_\mathrm{AT-all}$ = 0.4 compared to $\beta_\mathrm{GC-end}$ = 0.7, $\beta_\mathrm{GC-core}$ = 0.6, and $\beta_\mathrm{GC-mix}$ = 0.6. We can interpret the stretch as fraying from many configurations, each having a slightly different rate. The greater fraying heterogeneity is in agreement with the diverse kinetic landscape of the AT-all MSM.

@@

\noteb{check out the options for TOC}
\section*{TOC Image}

\begin{figure}[ht!]
	\begin{center} 
        \includegraphics[width=150mm, scale=1]{TOC_alternate2.pdf}
        %\caption*{TOC image}
        \label{fig:TOC}
	\end{center}
\end{figure}

>>>>

%include hyperlinks to figures
%\documentclass{article}
\usepackage{graphicx}
\usepackage[%  
    colorlinks=true,
    pdfborder={0 0 0},
    linkcolor=black,
    citecolor=black,
    urlcolor=black
]{hyperref}

@@

\begin{figure}[ht!]
	\begin{center}
        \includegraphics[width=1.0\textwidth]{Fig2.pdf}
	\end{center}
\end{figure}
\begin{figure}[ht!]
        \caption{Thermodynamic and kinetic predictions of the sequence-specific MSMs for AT-all, GC-end, GC-core, and GC-mix. (a) \textbf{Macrostates.} Schematic representation of the seven metastable macrostates occupied by one or more of the four sequences: fully hybridized state (H), two-base out-of-register shifted states in the 5$^\prime$ (5S2) or 3$^\prime$ (3S2) direction, four-base out-of-register shifted states in the 5$^\prime$ (5S4) or 3$^\prime$ (3S4) direction, and the fully dissociated state (D). The line drawings represent the 10-base self-complementary sequences, where red-to-blue contacts indicates (possible) WC base pairing and black indicates an unbound bases. Adjacent to each line drawing we provide representative molecular structures corresponding to that macrostate. (b) \textbf{Thermodynamics.} Histograms reporting the number of the $10^7$ total frames within the 1 ms of simulation trajectories observed to occupy each of the seven macrostates, corresponding to our numerical estimates of the equilibrium occupancy probabilities. Uncertainties are calculated across 100 MSMs using a Bayesian MSM estimation are reported for each bar and are very small compared to the total counts. (c) \textbf{Kinetic networks.} MSMs illustrating the kinetic network for each sequence. The orange circles correspond to the macrostates occupied by each sequence and are labeled by the macrostate codes reported in panel a. The radii of the circles is proportional to the logarithm of the equilibrium occupancy probabilities reported in panel b. Molecular renderings of an illustrative snapshot from the coarse-grained molecular simulations are provided next to each macrostate. The grey arrows between macrostates indicate the presence of a probability flux between this pair of states at equilibrium and the arrow thickness is proportional to the flux. (To avoid congesting the diagram, arrows are not reported for probability fluxes lower than $3\times10^{-6}$.) The numerical value overlaid on each arrow reports the conditional probability that a system occupying the macrostate at the start of the arrow at time $t$ will transition to the macrostate at the end of the arrow by time $(t+\tau)$, where $\tau$=1.2 ns is the lag time corresponding to a single time step of the MSM. Large orange circles correspond to thermodynamically favorable states and large grey arrows correspond to kinetically favorable transitions. (d) \textbf{Kinetic time scales.} Distribution of MSM implied time scales for each sequence. The leading implied time scale corresponds to the characteristic time scale for hybridization/dehybridization and is approximately equal for all systems since simulations were conducted at the same concentration and at the respective melting temperatures. The higher order implied time scales correspond to a spectrum of kinetic relaxations between the constituent macrostates in the MSM corresponding to shifted and/or frayed states.
    }
    \label{fig:allseq_table}
\end{figure}

@@

AT-all possesses the richest and most complex MSM of the four sequences by virtue of its repetitive nature, comprising a hybridized state (H), dissociated state (D), and four out-of-register shifted states (5S2, 3S2, 5S4, 3S4) (Fig.~\ref{fig:allseq_table}c). Analysis of the MSM transition probabilities reveal a critical role of the shifted states in mediating hybridization and dehybridization. Commencing from the dissociated state D, we observe approximately equal probabilities for transitions to each of the other five states, such that a transition to one of the out-of-register shifted states 5S2, 3S2, 5S4, or 3S4 is approximately 2.2 times more likely than a direct transition to the hybridized state H. Commencing from the hybridized state H, however, a direct transition to the dissociated state is approximately 1.2 times more likely than a transition to one of the two-base shifted states 5S2 or 3S2. Once in one of the four shifted states, the 5$^\prime$ vs. 3$^\prime$ overhang and degree of shifting play an important role in determining whether the duplex will transition to more shifted states, more aligned states, or completely dissociate. Transitions from more shifted states towards more aligned states (i.e., 5S4 $\rightarrow$ 5S2, 5S2 $\rightarrow$ H, 3S4 $\rightarrow$ 3S2, 3S2 $\rightarrow$ H) are approximately an order of magnitude more probable than the reverse transitions from more aligned states to more shifted states. The largest single transition probability from the four shifted states 5S2, 3S2, 5S4, and 3S4 is, however, back to the dissociated state D. Consistent with the higher destabilizing effect of 3$^\prime$ inert tails relative to 5$^\prime$ tails,~\citep{Doktycz1990ThermodynamicATGC, Dickman2012ThermodynamicDNAs, Michele2014EHybridization} the 3S4 $\rightarrow$ D transition probability is twice as large as the 5S4 $\rightarrow$ D, and the 3S2 $\rightarrow$ D is four times larger than the 5S2 $\rightarrow$ D. The transition probability from the 5S2 and 3S2 states back to the dissociated state D is equal to or greater than the transition probability to the hybridized state. A transition path theory analysis of the MSM reveals that 33\% of productive hybridization trajectories D $\dashrightarrow$ H (where the dashed arrow indicates the combination of both direct and indirect pathways) proceed through one or more out-of-register shifted states. Among these out-of-register pathways, the D $\dashrightarrow$ 5S2 $\dashrightarrow$ H transition is predicted to occur 57\% of the time. A mean first passage time (MFPT) analysis returns a MFPT for D $\dashrightarrow$ H of 3.0 $\mu$s and for H $\dashrightarrow$ D of 2.5 $\mu$s. As expected by the fact that the calculations are performed at the approximate melting temperature of the sequence, the MFPTs are approximately equal to one another.

%\noteb{Can add additional analysis "When considering the mean MFPT between D and any state A (A = 5S2 $\cup$ 3S2 $\cup$ 5S4 $\cup$ 3S4 $\cup$ H), we observe the shorter time scales 1.3 $\mu$s for A $\rightarrow$ D 1.3 $\mu$s for D $\rightarrow$ A. These MFPTs reflect the frequency of partial hybridization events and are in close agreement with the leading implied time scale of $t_1$ = 1.5 $\mu$s reported in Fig.~\ref{fig:allseq_table}d" This works for AT-all, but for all sequences the leading timescale tends to be lower by up to 2x than H $\dashrightarrow$ D MFTPs. I suspect this is because these leading timescale takes longer to converge, but I'm not sure if we want to draw attention to this and/or report the leading timescale for each sequence since its exact meaning is not always clear)}

@@

%\noteb{and in good agreement with the leading implied time scale of $t_1$ = 1.4 $\mu$s reported in Fig.~\ref{fig:allseq_table}d.}

@@

%\noteb{This MFPT shows the largest deviation from the leading timescale $t_1$ = 1.6 $\mu$s}

@@

%\noteb{Again, these are nearly equal and commensurate with the dominant leading implied time scale of $t_1$ = 2.0 $\mu$s reported in Fig.~\ref{fig:allseq_table}d. }

@@

\subsection{\label{sec:stretch_comparison}Long-lived metastable shifted states predicted by the MSM are resolved by T-jump IR}

Finally, we sought to validate the predictions of our sequence-dependent MSMs against experimental T-jump IR spectroscopy. T-jump IR measurements commence from a low temperatures, applies a step jump in temperature, and tracks the relaxation of the system to the dehybridized state. We hypothesized that the influence of the out-of-register shifted states present in the AT-all and GC-end sequences upon the system relaxation kinetics should be manifest in the slow and/or fast responses measured by T-jump IR. As discussed in Section \ref{sec:time}, the slow IR response corresponds to dissociation and the fast to terminal base fraying. With regards to the slow response, our MFPT analyses of our MSMs predict out-of-register shifting events (i.e., H $\rightarrow$ 3S2, 5S2, 3S4, 5S4) to proceed on \rood{microsecond} time scales, which are commensurate with the $\sim$\rood{XX} $\mu$s time scales for dehybridization (i.e., H $\rightarrow$ D). As such, we anticipate that the dynamical relaxations associated with out-of-register shifted states proceed on similar time scales to, and may not be distinguishable from, the relaxation to the dehybridized state. Nevertheless, the presence of these out-of-register shifted states in the low-temperature equilibrium ensemble prior to the T-jump step may be observable via their influence on the fast T-jump IR response attributable to fraying. Specifically, we hypothesize that the dangling ends and inert tails present in the out-of-register shifted states should promote an elevated fraying response over the course of the temperature relaxation, and should therefore be manifest in the observation of a \rood{stronger / slower / longer-lived} fast fraying response corresponding to a \rood{more stretched relaxation over \rood{XX}-\rood{XX} ns time scales}. Analysis of the MSM equilibrium distributions (Fig.~\ref{fig:allseq_table}b) reveals 24\% of the equilibrium ensemble to reside in out-of-register shifted states 3S2, 5S2, 3S4, and 5S4 for AT-all, compared to just 0.3\% for GC-end, and 0\% for GC-core and GC-mix. In is our conjecture that the substantial presence of out-of-register shifted states in the pre-T-jump AT-all ensemble should be distinguishable from the GC-end, GC-core, and GC-mix as an \rood{elongation of the fast relaxation response associated with terminal base fraying}.
% We can interpret the stretch as fraying from many configurations, each having a slightly different rate.

We present in Fig.~\ref{fig:T-jump_stretching} our T-jump IR t-HDVE difference spectra and corresponding normalized time traces at 1600 cm$^{-1}$ and 1660 cm$^{-1}$. The signal at 1600 cm$^{-1}$ corresponds to changes in A and T ring vibrations while the signal at 1660 cm$^{-1}$ contains contributions from G and T carbonyl vibrations. Each time trace was fitted to the sum of a stretched exponential and two exponentials $S(t) = A\exp{(-(t/\tau^{\mathrm{fast}})^{\beta})} + B\exp{(-t/\tau_{\mathrm{slow}})}+C\exp{(-t/\tau_{\mathrm{cool}})}$. The stretched exponential describes the relaxation process from 5 ns to 1 $\mu$s, and the two exponentials describe the signal increase from 1-320 $\mu$s and signal decay from re-hybridization induced by thermal relaxation back to the initial temperature. The fitting parameters $A$, $B$, and $C$ correspond to the relative amplitudes of the three kinetic responses and the stretch factor $\beta$ to the \rood{heterogeneity of fraying dynamics at the fast timescale} \noter{Whatever you write here should be consistent with the text in the hypothesis paragraph above.}. A testable prediction of our hypothesis is that the long-lived out-of-register shifted states in AT-all should result in a significantly smaller value for the fitted $\beta$ parameter relative to those for GC-end, GC-core, and GC-mix. This hypothesis was supported by the experimental time series at both 1600 cm$^{-1}$, where $\beta_\mathrm{AT-all}$ = 0.3 compared to $\beta_\mathrm{GC-end}$ = 0.6, $\beta_\mathrm{GC-core}$ = 0.7, and $\beta_\mathrm{GC-mix}$ = 0.6, and 1660 cm$^{-1}$, where $\beta_\mathrm{AT-all}$ = 0.4 compared to $\beta_\mathrm{GC-end}$ = 0.7, $\beta_\mathrm{GC-core}$ = 0.6, and $\beta_\mathrm{GC-mix}$ = 0.6. This result supports our hypothesis and validates a testable dynamical prediction of our sequence-dependent MSMs.

@@

\section*{TOC Image}

\begin{figure}[ht!]
	\begin{center} 
        \includegraphics[width=150mm, scale=1]{TOC.pdf}
        %\caption*{TOC image}
        \label{fig:TOC}
	\end{center}
\end{figure}
